Automatically Splitting a Two-Stage Lambda Calculus

### 2015-10-14 paper/relatedwork.tex
changes all over

In an attempt to reduce the complexity of the specialization phase of
self-applicable partial evaluators, researchers explored the use of
staging transformations that, much like splitting, partition an input
program into two programs: one that produces static outputs from
static inputs, and a second that produces dynamic outputs from both
static and dynamic inputs.  These metastatic transformations, referred
to as \emph{program bifurcation}\,\cite{DeNiel91} or as \emph{binding
  time separation}\,\cite{Mogensen89a}), serve as a preprocessing step
to specialization and were successfully applied to functional
languages that supported recursion\,\cite{DeNiel91}. However, unlike
splitting, where the goal is to emit code where first stage results
are computed once and then reused in multiple invocations of second
stage execution, the second function produced by program bifurcation
does not use the results of the first; Instead,
%only serves as a preprocessing step for the subsequent
%specialization phase of partial evaluation.
it has access to the first stage inputs and recomputes all required
first stage computations. (As noted by Knoblock and
Ruf\,\cite{knoblock96}, it may be possible to modify the program
bifurcation algorithm to cache and reuse the intermediate results, but
this was never attempted.) The grammar-based binding-time
specifications used in program bifurcation were capable of describing
data structures with purely-static, purely-dynamic, and mixed stage
content, much like the type system of \lang.

>>>>

Researchers have explored other staging transformations that, 
like splitting, partition an input program into static and dynamic components.  
In particular, \emph{binding time separation}\,\cite{Mogensen89a} 
(also called \emph{program bifurcation}\,\cite{DeNiel91})
has been used as a preprocessor step in partial evaluators,
allowing more efficient specialization of programs with 
mixed-stage data structures without changes to the specializer itself.
Notably, the grammar-based binding-time specifications used in binding 
time separation were capable of describing
data structures with purely-static, purely-dynamic, and mixed stage
content, much like the type system of \lang\
(though this correspondence is less clear without our addition of 
\bbonep\ and $\curr$).

That said, unlike splitting, where the goal is to emit code where first stage results
are computed once and then reused in multiple invocations of second
stage execution, the second (dynamic) function produced by binding time separation
does not use the results of the first; instead,
%only serves as a preprocessing step for the subsequent
%specialization phase of partial evaluation.
it has access to the first stage inputs and recomputes all required
first stage computations. As noted by Knoblock and
Ruf\,\cite{knoblock96}, it may be possible to modify the program
bifurcation algorithm to cache and reuse the intermediate results, but
this was never attempted.

### 2015-10-14 paper/splitting.tex
changes all over

In the recursive case, where the evaluation takes at least one step, we have 
$e \stepsym \bbtwo e' \stepsym \bbtwo \cdots \stepsym \bbtwo \exv q$
as well as \mbox{$\typestwo [\cdot] e A$} 
and \mbox{$\splittwo e A p l r$}.
By preservation and totality of splitting, we know $\typestwo [\cdot] {e'} A$ and $\splittwo {e'} A {p'} {l'} {r'}$.
From here, the inductive hypothesis yields $\reduce {p'} {b'}$ and $[b'/l']r' \equiv q$.
All that we now require is $\reduce p b$ and $[b/l]r \equiv [b'/l']r'$,
To close this gap we introduce \ref{lem:twoCorrect}, 
which essentially states that any single step is correct,
and whose proof will concern the rest of this section.

\begin{lemma} [Single Step Correctness]
\label{lem:twoCorrect}
\begin{itemize} 
\item If $e \stepsym \bbtwo e'$, $\splittwo e A p  l r$, $\splittwo {e'} A {p'} {l'} {r'}$, and $\reduce {p'} {b'}$,
then $\reduce p b$ and $[b/l]r \equiv [b'/l']r'$.
\item If $e \stepsym \bbonem e'$, $\splitone e A c {l.r}$, $\splitone {e'} A {c'} {l'.r'}$, and $\reduce {c'} {\tup{i,b'}}$,
then $\reduce c {\tup {i,b}}$ and $[b/l]r \equiv [b'/l']r'$.
\end{itemize}
\end{lemma}

After invocation of that new lemma, we can derive $[b/l]r \equiv [b'/l']r' \equiv q$ directly.
In order to prove \ref{lem:twoCorrect},
we will need to state analogous version for steps at \bbonem,
since the various kinds of multistage term in \lang\ are mutually dependent.
Thus, this section proceeds by covering the definition of splitting at \bbonem, 
starting with the value form at that world.

@@

The correctness of \bbonem\ term splitting with respect to a $\stepsym \bbonem$ step is given in \ref{lem:oneCorrect}.
It's very similar to the world \bbtwo\ version, 
in that the reduction of the first stage part of $e'$ should imply reduction of the first stage part of $e$,
and that the resulting applied resumers should be equivalent.
But split forms at \bbonem\ have one more piece of output than those at \bbtwo,
namely the immediate value $i$.
\ref{lem:oneCorrect} accounts for this by saying that 
that immediate value must be exactly identical on both sides of the step.

@@

In this section, we consider a few exemplar cases 
from the proofs of \ref{lem:oneCorrect} and \ref{lem:twoCorrect}.
The equivalence relation ``$\equiv$'' is defined in \ref{fig:equiv}.

@@

\paragraph{Compatibility of {\tt pi1} at \bbtwo.}
We are given 
$\steptwo {\pio e} {\pio {e'}}$, 
$\splittwo {\pio e} A p l {\pio r}$, 
$\splittwo {\pio {e'}} A {p'} {l'} {\pio {r'}}$,
and $\reduce {p'} {b'}$.
Inversion of the first three yields
$\steptwo e {e'}$, $\splittwo e A p l r$, and $\splittwo {e'} A {p'} {l'} {r'}$.
Using \ref{lem:twoCorrect} inductively gives $\reduce p b$ and $[b/l]r \equiv [b'/l']r$.
From this, $[b/l]\pio r \equiv [b'/l']\pio {r}$ can be derived directly.
This pattern generalizes to all world \bbtwo\ compatibility rules.

\paragraph{Compatibility of {\tt pi1} at \bbonem.}
We are given
$\stepmix {\pio e} {\pio {e'}}$, 
$\splitone {\pio e} A {\letin{\tup{y,z}}{c}{\tup{\pio {\exv y},\exv z}}} {l.\pio r}$, \\
\mbox{$\splitone {\pio {e'}} A {\letin{\tup{y,z}}{c'}{\tup{\pio {\exv y},\exv z}}} {l'.\pio {r'}}$},
and \\ $\reduce {\letin{\tup{y,z}}{c'}{\tup{\pio {\exv y},\exv z}}} {\tup{i,b'}}$.
Inversion of the first three yields 
$\stepmix e {e'}$, $\splitone e A c {l.r}$, and $\splitone {e'} A {c'} {l'.r'}$, 
and inversion of the reduction yields $\reduce {c'} {\tup{\tup{i,i_2},b'}}$.
Using \ref{lem:oneCorrect} inductively gives $\reduce c {\tup{\tup{i,i_2},b}}$ and $[b/l]r \equiv [b'/l']r$.
From this, we can derive $\reduce {\letin{\tup{y,z}}{c}{\tup{\pio {\exv y},\exv z}}} {\tup{i,b}}$ and 
$[b/l]\pio r \equiv [b'/l']\pio {r}$ directly.
This pattern generalizes to all world \bbonem\ compatibility rules.

>>>>

In the recursive case, where the evaluation takes at least one step, we have 
$e \stepsym \bbtwo e' \stepsym \bbtwo \cdots \stepsym \bbtwo \exv q$
as well as \mbox{$\typestwo [\cdot] e A$} 
and \mbox{$\splittwo e A p l r$}.
By preservation and totality of splitting, we know $\typestwo [\cdot] {e'} A$ and $\splittwo {e'} A {p'} {l'} {r'}$.
From here, the inductive hypothesis yields $\reduce {p'} {b'}$ and $[b'/l']r' \equiv q$.
All that we now require is $\reduce p b$ and $[b/l]r \equiv [b'/l']r'$,
To close this gap we introduce \ref{lem:stepCorrect}, 
which essentially states that any single step is correct,
and whose proof will concern the rest of this section.

\begin{lemma} [Single Step Correctness]
\label{lem:stepCorrect}
\begin{itemize} 
\item If $e \stepsym \bbtwo e'$, $\splittwo e A p  l r$, $\splittwo {e'} A {p'} {l'} {r'}$, and $\reduce {p'} {b'}$,
then $\reduce p b$ and $[b/l]r \equiv [b'/l']r'$.
\item If $e \stepsym \bbonem e'$, $\splitone e A c {l.r}$, $\splitone {e'} A {c'} {l'.r'}$, and $\reduce {c'} {\tup{i,b'}}$,
then $\reduce c {\tup {i,b}}$ and $[b/l]r \equiv [b'/l']r'$.
\end{itemize}
\end{lemma}

After invocation of that new lemma, we can derive $[b/l]r \equiv [b'/l']r' \equiv q$ directly.
In order to prove \ref{lem:stepCorrect},
we will need to state analogous version for steps at \bbonem,
since the various kinds of multistage term in \lang\ are mutually dependent.
Thus, this section proceeds by covering the definition of splitting at \bbonem, 
starting with the value form at that world.

@@

The correctness of \bbonem\ term splitting with respect to a $\stepsym \bbonem$ step is given in \ref{lem:stepCorrect}.
It's very similar to the world \bbtwo\ version, 
in that the reduction of the first stage part of $e'$ should imply reduction of the first stage part of $e$,
and that the resulting applied resumers should be equivalent.
But split forms at \bbonem\ have one more piece of output than those at \bbtwo,
namely the immediate value $i$.
\ref{lem:stepCorrect} accounts for this by saying that 
that immediate value must be exactly identical on both sides of the step.

@@

In this section, we consider a few exemplar cases 
from the proof of \ref{lem:stepCorrect}.
The equivalence relation ``$\equiv$'' is defined in \ref{fig:equiv}.

@@

\paragraph{Compatibility of {\tt pi1} at \bbtwo.}
We are given 
$\steptwo {\pio e} {\pio {e'}}$, 
$\splittwo {\pio e} A p l {\pio r}$, 
$\splittwo {\pio {e'}} A {p'} {l'} {\pio {r'}}$,
and $\reduce {p'} {b'}$.
Inversion of the first three yields
$\steptwo e {e'}$, $\splittwo e A p l r$, and $\splittwo {e'} A {p'} {l'} {r'}$.
Using \ref{lem:stepCorrect} inductively gives $\reduce p b$ and $[b/l]r \equiv [b'/l']r$.
From this, $[b/l]\pio r \equiv [b'/l']\pio {r}$ can be derived directly.
This pattern generalizes to all world \bbtwo\ compatibility rules.

\paragraph{Compatibility of {\tt pi1} at \bbonem.}
We are given
$\stepmix {\pio e} {\pio {e'}}$, 
$\splitone {\pio e} A {\letin{\tup{y,z}}{c}{\tup{\pio {\exv y},\exv z}}} {l.\pio r}$, \\
\mbox{$\splitone {\pio {e'}} A {\letin{\tup{y,z}}{c'}{\tup{\pio {\exv y},\exv z}}} {l'.\pio {r'}}$},
and \\ $\reduce {\letin{\tup{y,z}}{c'}{\tup{\pio {\exv y},\exv z}}} {\tup{i,b'}}$.
Inversion of the first three yields 
$\stepmix e {e'}$, $\splitone e A c {l.r}$, and $\splitone {e'} A {c'} {l'.r'}$, 
and inversion of the reduction yields $\reduce {c'} {\tup{\tup{i,i_2},b'}}$.
Using \ref{lem:stepCorrect} inductively gives $\reduce c {\tup{\tup{i,i_2},b}}$ and $[b/l]r \equiv [b'/l']r$.
From this, we can derive $\reduce {\letin{\tup{y,z}}{c}{\tup{\pio {\exv y},\exv z}}} {\tup{i,b}}$ and 
$[b/l]\pio r \equiv [b'/l']\pio {r}$ directly.
This pattern generalizes to all world \bbonem\ compatibility rules.
